Individual Variation in Animal Movement and Effects on Seed Dispersal Distance

### 2021-01-15 paper/paper.Rmd
delete the tables
Updating with figures in the right place

---
title: "Individual Variation in Animal Movement and Effects on Seed Dispersal Distance"
abstract: |
  I don't like this abstract. Revisit at the end...Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an araçari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual araçari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output:
  html_document:
    df_print: paged
  word_document:
    reference_docx: paper_ref.docx
bibliography: mybibfile.bib
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

@@

\newpage
# Introduction
A main goal of ecologists and evolutionary biologists is to document and predict the structure and dynamics of ecological systems. Although most of our current understanding of ecological systems is based on the average patterns observed in nature, understanding variability and its consequences in ecological processes has recently raised the interests of researchers [@benedetti-cecchi_importance_2003; @inouye_importance_2005]. Variation is now considered a property of ecological systems that might contain as much ecological information as do averages and regularities in nature [@benedetti-cecchi_importance_2003; @violle_return_2012].
Individual variation, i.e. the possibility for individuals within a population to present different ecological strategies or traits [@violle_return_2012], has been widely discussed in ecology and evolutionary biology. In evolutionary biology the acknowledgement that individuals with different traits might vary in their fitness and survival under natural selection pressures dates back to Darwin’s seminal ideas [@darwin_origin_1859]. Today, it is widely recognized that individual variation plays a key role in the evolution of species and clades, and the relevance of individual variation in the performance of populations and species facing current threats (e.g. climate change, fragmentation and habitat loss) is a growing field of research [@forsman_inter-individual_2016]. In ecology, individual variation is often mentioned as a mechanism for populations to reduce competition pressure [@bolnick_ecological_2010; @araujo_ecological_2011]. Based on the niche complementarity principle, phenotypically different individuals will compete less strongly than more similar individuals [@bolnick_why_2011], thus individual variation would act as a release from intra- and interspecific competition at a local scale. Despite the well recognized importance of individual variation in ecological and evolutionary processes, ecological theory has emphasized interspecific variation [@violle_return_2012]. Not surprisingly, the role of individual variability on the outcome of species interactions has also been poorly studied; most models of species interactions assume that all conspecific individuals are equivalent and interchangeable [@bolnick_why_2011]. Individuals, however, differ in several aspects that will affect their interactions with other species. This variability makes individuals non-interchangeable from the perspective of interacting patterns [@bolnick_ecology_2003; @gonzalez-varo_labile_2016]. The consequences of individual variability in interspecies interactions can even have cascading effects, going beyond from the species-species level to the community and ecosystem level [@post_intraspecific_2008; @wolf_animal_2012].

@@

At the beginning of each simulation run, the animal consumed five seeds at the focus tree location, and each of these seeds was assigned a gut retention time sampled from a gamma distribution (shape=4, scale=5, shift=8). Once the animal in the simulation consumed the seeds, a movement distance was sampled from the assigned exponential distribution (which varied depending on the underlying model such as the null, individual or family group model) and a random direction from $0^o$ to $360^o$ degrees was used to determine the animal's position for the next time step, repeating this process for every time step and thus following an uncorrelated random walk with no directional tendency (*Edit references* Turchin 1998). Once the simulation run's time matched the gut retention time for a seed, that seed would get dropped at the animal's location at that specific time point, thus allowing us to record seed location in the landscape. For each of our models we ran 10,000 simulation runs per individual or family group, depending on the model, and collected information on animal and seed locations at every time step.

@@

@@

Where the shape parameter $a$ determines the tail of the distribution, and $b$ is the scale parameter. It is worth mentioning that the standard parameterization of the Weibull in 'R' is different from the parameterization used in Morales and Carlo  [-@morales_effects_2006], where the shape parameter $a = \nu$ remains the same, but the scale parameter is defined as $\kappa = b^{-\nu}$, giving a density distribution function of $W(r)= \kappa\nu r^{\nu-1} exp[-{\kappa r}^\nu]$. 
<!-- Here, we report the scale parameter following the alternative parameterization, but the original values returned after fitting the Weibull distribution using R can be found in the Supplementary Table 1.  -->
Since the shape parameter remains the same, we can note that for $a=\nu=1$ the distribution shows a tail with exponential decay, with values of $\nu <1$ the tail shows fast-decay, and when $\nu>1$ we can see a fat-tailed distribution (Fig 2.) We visually assessed the fit of the distributions via qqplots and used the Kolmogorov-Smirnov statistic to assess goodness of fit (Supplementary information).

@@

```{r}
knitr::kable(dispersion_table)
```

@@

**Table 1.**
```{r}

knitr::kable(dispersal_kernel_table)
```

>>>>

---
title: "Individual Variation in Animal Movement and Effects on Seed Dispersal Distance"
abstract: |
  I don't like this abstract. Revisit at the end...Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an araçari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual araçari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output:
  word_document:
    reference_docx: paper_ref.docx
  html_document:
    df_print: paged
bibliography: mybibfile.bib
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)
```

@@

```
\newpage
# Introduction
A main goal of ecologists and evolutionary biologists is to document and predict the structure and dynamics of ecological systems. Although most of our current understanding of ecological systems is based on the average patterns observed in nature, understanding variability and its consequences in ecological processes has recently raised the interests of researchers [@benedetti-cecchi_importance_2003; @inouye_importance_2005]. Variation is now considered a property of ecological systems that might contain as much ecological information as do averages and regularities in nature [@benedetti-cecchi_importance_2003; @violle_return_2012].
Individual variation, i.e. the possibility for individuals within a population to present different ecological strategies or traits [@violle_return_2012], has been widely discussed in ecology and evolutionary biology. In evolutionary biology the acknowledgement that individuals with different traits might vary in their fitness and survival under natural selection pressures dates back to Darwin’s seminal ideas [@darwin_origin_1859]. Today, it is widely recognized that individual variation plays a key role in the evolution of species and clades, and the relevance of individual variation in the performance of populations and species facing current threats (e.g. climate change, fragmentation and habitat loss) is a growing field of research [@forsman_inter-individual_2016]. In ecology, individual variation is often mentioned as a mechanism for populations to reduce competition pressure [@bolnick_ecological_2010; @araujo_ecological_2011]. Based on the niche complementarity principle, phenotypically different individuals will compete less strongly than more similar individuals [@bolnick_why_2011], thus individual variation would act as a release from intra- and interspecific competition at a local scale. Despite the well recognized importance of individual variation in ecological and evolutionary processes, ecological theory has emphasized interspecific variation [@violle_return_2012]. Not surprisingly, the role of individual variability on the outcome of species interactions has also been poorly studied; most models of species interactions assume that all conspecific individuals are equivalent and interchangeable [@bolnick_why_2011]. Individuals, however, differ in several aspects that will affect their interactions with other species. This variability makes individuals non-interchangeable from the perspective of interacting patterns [@bolnick_ecology_2003; @gonzalez-varo_labile_2016]. The consequences of individual variability in interspecies interactions can even have cascading effects, going beyond from the species-species level to the community and ecosystem level [@post_intraspecific_2008; @wolf_animal_2012].

@@

At the beginning of each simulation run, the animal consumed five seeds at the focus tree location, and each of these seeds was assigned a gut retention time sampled from a gamma distribution (shape=4, scale=5, shift=8, Figure 1a). Once the animal in the simulation consumed the seeds, a movement distance was sampled from the assigned exponential distribution (Figure 1b), which varied depending on the underlying model such as the null, individual or family group model. A random direction from $0^o$ to $360^o$ degrees was used to determine the angle of movement, which with the movement distance would determine the animal's position for the next time step, repeating this process for every time step and thus following an uncorrelated random walk with no directional tendency (*Edit references* Turchin 1998). Once the simulation run's time matched the gut retention time for a seed, that seed would get dropped at the animal's location at that specific time point, thus allowing us to record seed location in the landscape (Figure 2a). For each of our models we ran 10,000 simulation runs per individual or family group, depending on the model, and collected information on animal and seed locations at every time step.

```{r}
# Simulate the data for one run

set.seed(5)

data(ptpl)

# Panel 2: Table with gut retention times and the gamma distribution used for it
n <- 10000
grt_dens <- data.frame(seq = 1:n, dat = (rgamma(n, shape = 4, scale = 5) + 8))
grt_data <- data.frame(time = round(sample(grt_dens$dat, 5)), seq = 1:5)
tt <- max(grt_data$time)
ggplot(grt_dens, aes(x = dat)) +
  geom_density() + 
  geom_point(data = grt_data, aes(x = time, y = 0), color = "red") +
  #geom_point(aes(x = tt, y = 0), color = "red") +
  theme_bw() + labs(x = "Time in minutes", y = "Density") -> fig_GRT

# Panel 3: Exponential distribution used for animal movement lengths
rate <- 1/mean(ptpl$mpm)
md_dens <- data.frame(seq = 1:n, dat = rexp(n, rate = rate))
md_hist <- data.frame(md = round(sample(md_dens$dat, tt)))

ggplot(md_dens, aes(x = dat)) +
  geom_density() + 
  geom_point(data = md_hist, aes(x = md, y = 0)) +
  theme_bw() + labs(x = "Distance in meters", y = "Density") -> fig_MD

ggplot(md_hist, aes(x = md)) +
  geom_histogram() +
  theme_bw() + labs(x = "Distance in meters", y = "Count") -> fig_MD_hist

# Panel 4: Animal trajectory and locations of the seeds at the time the seeds are dropped
set.seed(5)
distance <- md_hist$md
angle <- runif(tt, min = 0, max = 360)
distx <- distance*cos(angle)
xloc <- c(0, cumsum(distx))
disty <- distance*sin(angle)
yloc <- c(0, cumsum(disty))
animalTraj <- data.frame(time = 0:tt, xloc = xloc, yloc = yloc)
seed_loc <- merge(animalTraj, grt_data, by = "time")
mean_seed_loc <- data.frame(x = mean(seed_loc$xloc), y = mean(seed_loc$yloc))

ggplot() +
  geom_path(data = animalTraj, aes(x = xloc, y = yloc), size = 1, color = "darkgrey") +
  geom_point(data = animalTraj, aes(x = xloc, y = yloc), color = "black", size = 1) +
  geom_point(data = seed_loc, aes(x = xloc, y = yloc), color = "red", size = 3) +
  geom_point(aes(x = 0, y = 0), color = "black", size = 5)+
  geom_vline(aes(xintercept=0), linetype = "dotted") + geom_hline(aes(yintercept = 0), linetype = "dotted") +
  labs(x = "x", y = "y") +
  coord_fixed() +
  theme_bw() -> fig_traj

ggplot() +
  geom_point(data = seed_loc, aes(x = xloc, y = yloc), color = "red", size = 3) +
  geom_point(aes(x = 0, y = 0), color = "black", size = 5)+
  geom_vline(aes(xintercept=0), linetype = "dotted") + geom_hline(aes(yintercept = 0), linetype = "dotted") +
  geom_segment(aes(x = seed_loc$xloc, xend = 0, y = seed_loc$yloc, yend = 0), linetype = "dashed", color = "red") +
  labs(x = "x", y = "y") +
  coord_fixed() +
  theme_bw() -> fig_disp

ggplot() +
  geom_point(data = seed_loc, aes(x = xloc, y = yloc), color = "red", size = 3) +
  geom_point(aes(x = 0, y = 0), color = "black", size = 5)+
  geom_vline(aes(xintercept=0), linetype = "dotted") + geom_hline(aes(yintercept = 0), linetype = "dotted") +
  #geom_segment(aes(x = seed_loc$xloc, xend = 0, y = seed_loc$yloc, yend = 0), linetype = "dashed", color = "red") +
  geom_point(data = mean_seed_loc, aes(x = x, y = y), color = "blue", size = 4) +
  geom_segment(aes(x = seed_loc$xloc, xend = mean_seed_loc$x, y = seed_loc$yloc, yend = mean_seed_loc$y), linetype = "dashed", color = "blue") +
  labs(x = "x", y = "y") +
  coord_fixed() +
  theme_bw() -> fig_agg

#### Figure 1. 
Density distributions from which we sample gut retention time (GRT) and movement distance (MD).

```{r fig.width=2.5, fig.height=4}
plot_grid(fig_GRT, fig_MD, align = "v", nrow = 2, labels = "auto", label_size = 10)
```

@@

#### Figure 2.
Simulation example where animal movement determines location for seeds dropped. Calculations for seed dispersal distance from the parent tree and seed distance to the mean seed location as a measure of aggregation.

```{r fig.height=2, fig.width=6}
plot_grid(fig_traj, fig_disp, fig_agg, ncol = 3, align = "h", labels = "auto", label_size = 8)
```

@@

Where the shape parameter $a$ determines the tail of the distribution, and $b$ is the scale parameter. It is worth mentioning that the standard parameterization of the Weibull in 'R' is different from the parameterization used in Morales and Carlo  [-@morales_effects_2006], where the shape parameter $a = \nu$ remains the same, but the scale parameter is defined as $\kappa = b^{-\nu}$, giving a density distribution function of $W(r)= \kappa\nu r^{\nu-1} exp[-{\kappa r}^\nu]$. 
<!-- Here, we report the scale parameter following the alternative parameterization, but the original values returned after fitting the Weibull distribution using R can be found in the Supplementary Table 1.  -->
Since the shape parameter remains the same, we can note that for $a=\nu=1$ the distribution shows a tail with exponential decay, with values of $\nu <1$ the tail shows fast-decay, and when $\nu>1$ we can see a fat-tailed distribution (Fig 3.) We visually assessed the fit of the distributions via qqplots and used the Kolmogorov-Smirnov statistic to assess goodness of fit (Supplementary information).

#### Figure 3.
Variation in Weibull distributions determined by the value of the shape parameter. The scale parameter is defined as $b=1$, and the shape parameters are defined as follows: solid line $a=\nu=1$ shows an exponential decay tail, dashed line $a=\nu=0.2$ shows a thin tail, and $a=\nu=2$ describes a fat tail with the dotted line.

```{r fig.width=3, fig.height=2.5}
# shape
exptail <- 1
thintail <- 0.2
fattail <- 2

#scale
b <- 1
kappaexp <- b^-exptail
kappathin <- b^-thintail
kappafat <- b^-fattail

ggplot(data = data.frame(x = c(0.1, 4)), aes(x)) +
  stat_function(fun = dweibull, n = 100, args = list(scale = b, shape = thintail),
                color = "black", linetype = "dashed", size = 0.8) +
  stat_function(fun = dweibull, n = 100, args = list(scale = b, shape = exptail), 
                color = "black", linetype = "dotted", size = 0.8) +
  stat_function(fun = dweibull, n = 100, args = list(scale = b, shape = fattail),
                color = "black", linetype = "solid", size = 0.8) +
  ylab("") + xlab("") +
  theme_bw()

@@

@@
